\chapter{Boolean Satisfiability}

\section{Chaff}

\begin{enumerate}
\item \emph{Implication}: Propositional variable forced to be true by the current assignment of the
      propositional variables.
\item \emph{Decision}: Assignment of setting a variable to either \texttt{true} or \texttt{false}.
\item \emph{Decision Stack}: Stack of all decisions.
\item \emph{Decision Level}: Integer tag identifying the last \emph{decision}.
\item \emph{Decision Assignment}: Which variable is assigned next, and which value do we pick?
\item \emph{implied clause}: A clause is implied iff all but one of the literals of the clause have
      a value of \texttt{true} given the current variable assignment.
\end{enumerate}

\subsection{Watch Lists}
In a typical DP \textsc{Sat} solver, $90\symbol{37}$ of the computation time is spend in Boolean
constraint propagation.  Therefore, it is essential to get Boolean constraint propagation fast.
 
In every clause, two literals are \emph{watched}.  If any of these two literals evaluates to \texttt{true}
because of a variable assignment, we do not need to watch the clause any more, as it evaluates as 
\texttt{true}.  If one of these variables is assigned \texttt{false}, there are two cases:
\begin{enumerate}
\item If all the unwatched literals in the variable have been assigned 
      \texttt{false}, the remaining watched literal is assigned \texttt{true}.
\item If at least one of the unwatched literals in the variable have been assigned 
      \texttt{true}, the clause is \texttt{true} and does not need to be watched anymore.
\item If there is an unwatched literal that is not yet assigned, we watch this literal instead of
      the literal that evaluates as \texttt{false} now.
\end{enumerate}
The choice of which variables to watch is not important.  If a variable is unassigned, we do not
need to change the watch lists.

Data Structure: A map that maps propositional variables to lists of clauses.

\subsection{Decaying Sum Heuristic}
\begin{enumerate}
\item Each literal has a counter that is initialized to $0$.
\item When a clause is added to the database, the counter of each literal occurring in the clause
      is incremented.
\item Choose the unassigned variable with the highest count.
\item Break ties randomly.
\item Periodically, all counters are divided by a constant. (This is the decaying process.)
\end{enumerate}

\subsection{Restarts}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "artificial-intelligence"
%%% End:
